### 2021-04-02 newvariant.tex
Trying to clarify early explanations

\newcommand{\rx}[1]{\ensuremath{{r}_{#1}}\xspace} 
\newcommand{\ry}[1]{\rx{\mathrm{#1}}}

\newcommand{\Rx}[1]{\ensuremath{{\mathcal R}_{#1}}\xspace} 
\newcommand{\Ry}[1]{\Rx{\mathrm{#1}}}
\newcommand{\Ro}{\Rx{0}}
\newcommand{\Rc}{\Rx{\mathrm{c}}}
\newcommand{\Ri}{\Rx{\mathrm{i}}}
\newcommand{\RR}{\ensuremath{{\mathcal R}}\xspace}
\newcommand{\Rhat}{\ensuremath{{\hat\RR}}}
\newcommand{\Rnaive}{\ensuremath{{\mathcal R}_{\textrm{\tiny naive}}}\xspace}
\newcommand{\tsub}[2]{#1_{{\textrm{\tiny #2}}}}
\newcommand{\dd}[1]{\ensuremath{\, \mathrm{d}#1}}
\newcommand{\dtau}{\dd{\tau}}
\newcommand{\dx}{\dd{x}}
\newcommand{\dsigma}{\dd{\sigma}}

@@

Epidemic speed $r_x$ can be often estimated directly from case data \citep{mills2004transmissibility,nishiura2009transmission,ma2014estimating}.
However, estimating epidemic speed from case data can be challenging when case counts are low (e.g., in the beginning of the spread of the new variant) and is sensitive to temporal changes in testing patterns and intensity.
On the other hand, studies of new SARS-CoV-2 variants have mostly focused on characterizing changes in \emph{proportion} of a new variant \citep{switzerland2021variant, davies2021estimated, di2021impact, leung2021early, volz2021transmission,zhao2021}.
This has the advantage of being less sensitive to such changes in testing, and possibly for other transient effects that might affect variants and wild type viruses similarly.
When incidence is changing exponentially ($i_x(t) = i_x(t_0) \exp(r_x t)$), the proportion of the new variant $p(t)$ follows a logistic equation:
\begin{align}
p(t) &= \frac{\iy{var}(t_0) \exp(\ry{var} t)}{\iy{wt}(t_0) \exp(\ry{wt} t) + \iy{var}(t_0) \exp(\ry{var} t)},
\\ &= \frac{1}{1 + \left(\iy{wt}(t_0)/\iy{var}(t_0)\right) \exp(-\delta t)},
\end{align}
where the logistic growth rate $\delta = \ry{var} - \ry{wt}$ corresponds to the relative speed of the epidemic.

\begin{figure}[!th]
\includegraphics[width=\textwidth]{relstrength.pdf}
\caption{
\textbf{Relative strength of the new variant given observed relative speed.}
(A) Estimated relative strength of the new variant $\hat{\rho}$ given speed of the wild type
$r_w$ assuming identical generation-interval distributions.
(B) True relative strength of the new variant $\rho$ given speed of the wild type $r_w$ and ratio between the mean generation interval of the new variant $\bar{G}_v$ and that of the wild type $\bar{G}_w$.
(C) Ratio between the true and the estimated relative strength of the new variant $\rho/\hat{\rho}$.
True relative strength of the new variant $\rho$ is calculated using $\delta=0.1\,\textrm{days}$, $\bar{G}_w = 5\,\textrm{days}$, and $\kappa = 1/5$.
}
\label{fig:relstrength}
\end{figure}

As our baseline scenario, we assume $\delta = 0.1/\textrm{days}$ based on the observed relative speed of the new variant in the UK \citep{davies2021estimated}.
This speed can be then used to estimate the relative strength $\rho = \Ry{var}/\Ry{wt}$ of the new variant with a caveat; such inference depends on assumptions about the generation-interval distribution.
The mean generation interval $\bar{G}_w$ of the wild type is assumed to equal 5 days \citep{ferretti2020quantifying}.
The squared coefficient of variation $\kappa$ in generation intervals os assumed to equal 0.2 \citep{ferretti2020quantifying}---for simplicity, we assume that both the variant and the wild type have equal $\kappa$ and only consider differences in the mean.
As we expect there to be greater uncertainty in $\kappa$ than in $\bar{G}$, we evaluate the estimates of relative strength $\rho$ across a wide range of $\kappa$, ranging from 0 (delta distribution) to 1 (exponential distribution).

Previous studies have assumed that the generation-interval distributions do not differ between the wild type and the variant: $\Gy{var} = \Gy{wt} = \bar{G}$.
In this case, the estimate of the relative strength is given by \citep{park2019practical}:
\begin{equation}
\hat{\rho} = \left(1 + \frac{\kappa \delta \bar{G}}{1 + \kappa \ry{wt} \bar{G}}\right)^{1/\kappa}.
\end{equation}
Therefore, estimates of relative strength $\rho$ depends not only on the relative speed $\delta$ but also on how fast the wild type is spreading in the population $r_w$---a few analyses have implicitly or explicitly neglected this factor by either assuming $r_w = 0$ \citep{switzerland2021variant} or $\kappa = 0$ \citep{davies2021estimated}, in which case we get $\hat{\rho} = \exp(\delta \bar{G})$.
Instead, if the wild type was spreading faster than previously assumed (higher $r_w$), the estimates of the relative strength $\rho$ will be go down (\fref{relstrength}A).
The relative strength also depends on the \emph{effective dispersion} in generation intervals (i.e., the amount of variability in generation intervals, as measured by $\kappa$)---using narrower distributions makes $\hat{\rho}$ less sensitive to $\ry{wt}$.

However, if the true mean generation interval of the new variant is different from that of the wild type ($\bar{G}_v \neq \bar{G}_w$), the true relative strength $\rho = \RR_v/\RR_w$ is given by:
\begin{equation}
\rho = \left(\frac{1 + \kappa (r_w + \delta) \bar{G}_v}{1 + \kappa r_w \bar{G}_w}\right)^{1/\kappa}.
\end{equation}
In this case, longer (or shorter) mean generation interval of the new variant generally translates to higher (or lower) values of $\rho$---\fref{relstrength}B illustrates this effect assuming $\kappa = 1/5$.
However, this direction can change when both the new variant and the wild type are decaying exponentially (see area below $r_v = 0$ line in \fref{relstrength}B)---when disease incidence is decaying exponentially, longer generation intervals lead to a \emph{lower} value of strength (and therefore lower relative strength $\rho$).

Therefore, neglecting potential differences in generation intervals can lead to both under- and over-estimation of the relative strength.
In \fref{relstrength}C, we quantify the degrees of bias using the ratio between the true value $\rho$, which accounts for differences in generation-interval distributions, and the estimated value $\hat{\rho}$, which assumes equal generation-interval distributions.
Across the ranges of parameters that we consider, we find that up to 1.4 fold (either upward or downward) adjustments to current estimates may be required.

@@

While relative speed is easier to measure, we often do not expect $\delta$ to be constant---this can depend on epidemic and demographic factors and can vary across time and space.
Instead, biological mechanisms that cause new variant to have higher strength better translate to assuming a constant value of relative strength $\rho$.
For example, if the new variant has a higher transmission rate, the ratio between the strength of the new variant and that of the wild type will remain constant ($=\rho$) under strength-like interventions that reduces the strength of both strains by a constant amount.
However, this is not necessarily true under speed-like interventions (e.g., contact tracing and self-isolation), which depends on time since infection: if the generation-interval distributions are different, the effectiveness of interventions will also differ.
For now, we consider consequences of assuming a constant relative strength $\rho$ and predicting the relative speed $\delta$---an assumption often made (either implicitly or explicitly) in studies that have tried to predict the future spread of the new variant, including when it will replace the wild type.
As our baseline scenario we assume $\rho = 1.61$, which is the value we obtain for $\delta=0.1\,\textrm{days}$, $\bar{G}_w = \bar{G}_v = 5\,\textrm{days}$, and $\kappa = 1/5$.

\begin{figure}[!th]
\includegraphics[width=\textwidth]{relspeed.pdf}
\caption{
\textbf{Relative speed of the new variant given assumed relative strength.}
(A) Estimated relative speed of the new variant $\hat{\delta}$ given strength of the wild type $\RR_w$ assuming identical generation-interval distributions.
(B) True relative strength of the new variant $\delta$ given strength of the wild type $\RR_w$ and ratio between the mean generation interval of the new variant $\bar{G}_v$ and that of the wild type $\bar{G}_w$..
(C) Difference between the true and the estimated relative speed of the new variant $\delta - \hat{\delta}$.
True relative speed of the new variant $\delta$ is calculated using $\rho=1.61$, $\bar{G}_w = 5\,\textrm{days}$, and $\kappa=1/5$.
}
\label{fig:relspeed}
\end{figure}

@@

In practice, epidemiological conditions change over time due to intervention measures and susceptible depletion effects.
Such changes can be characterized by ``time-varying'' or ``instantaneous'' reproduction numbers, $\RR(t)$ \citep{fraser2007estimating}.
Assuming that the generation-interval distribution does not change over time, the instantaneous reproduction numbers of the new variant and of the wild type can be estimated from their corresponding incidence curves---an approach popularized by \cite{cori2013new}:
\begin{equation}
\RR_x(t) = \frac{i_x(t)}{\int_0^\infty i_x(t-\tau) g_x(\tau) \dtau}.
\end{equation}
Under strength-like intervention measures that reduce transmission rates by a constant amount, we expect ratios between true reproduction numbers to remain constant and correspond to the true relative strength: $\Ry{var}(t)/\Ry{wt}(t) = \rho$.
However, if the assumed generation-interval distribution $\hat{g}(\tau)$ is different from the true distribution, then the ratio between the estimated reproduction numbers $\hat{\rho}(t) = \hat{\RR}_{\textrm{var}}(t)/\hat{\RR}_{\textrm{wt}}(t)$ need not be constant, where
\begin{equation}
\hat{\RR}_x(t) = \frac{i_x(t)}{\int_0^\infty i_x(t-\tau) \hat{g}(\tau) \dtau}.
\end{equation}

Here, we simulate three different scenarios using a two strain renewal equation that assumes perfect cross-immunity (\fref{Rtbias}): 
(1) the wild type and the variant have the same generation-interval distributions, which match the assumed distribution (\fref{Rtbias}A--C);
(2) the assumed distribution matches the generation-interval distribution of the wild type but has a longer mean than the mean generation interval of the variant (\fref{Rtbias}D--F); and
(3) the assumed distribution matches the generation-interval distribution of the wild type but has a shorter mean than the mean generation interval of the variant (\fref{Rtbias}G--I).
Then, we compare the estimated ratio $\hat{\rho}(t) = \hat{\RR}_{\textrm{var}}(t)/\hat{\RR}_{\textrm{wt}}(t)$ with the true ratio $\rho = \Ry{var}(t)/\Ry{wt}(t)$.
In order to reflect smooth changes in $\RR$ driven by non-pharmaceutical interventions, we model $\Ry{wt}(t)$ phenomenologically and let $\Ry{var}(t) = \rho \Ry{wt}(t)$.

\begin{figure}[!th]
\includegraphics[width=\textwidth]{Rtbias.pdf}
\caption{
\textbf{Estimates of relative strength over time under different scenarios.}
(A, D, G) True (solid lines) and estimated (dashed lines) reproduction numbers of the new variant and the wild type over time.
(B, E, H) True (purple, solid) and estimated (orange, dashed) ratios between reproduction numbers of the new variant and the wild type over time.
(C, F, I) True (purple, solid) and estimated (orange, dashed) relationships between estimated reproduction numbers on a log-log sacle.
Top row: the assumed mean generation-interval (5 days) is equal to the mean generation-interval of the wild type and the variant.
Middle row: the assumed mean generation-interval (5 days) is equal to the mean generation-interval of the wild type but is longer than that of the varant (4 days).
Bottom row: the assumed mean generation-interval (5 days) is equal to the mean generation-interval of the wild type but is longer than that of the varant (6 days).
For all simulations, squared coefficient of variation in generation intervals is assumed to equal $\kappa = 1/5$.
We assume $\Ry{wt}(t)=2.5$ before $t=15\,\textrm{days}$ to reflect unmitigated spread and let $\Ry{wt}(t)=2.5 (1+0.7\cos((t-15)/10))/1.7$ after to model smooth introduction and lifting of non-pharmaceutical interventions.
}
\label{fig:Rtbias}
\end{figure}

When the assumed distribution matches the true distributions, the estimated reproduction numbers match the true values (\fref{Rtbias}A), and therefore the ratio remains constant (\fref{Rtbias}B).
However, if the generation-interval distribution of the variant is different from the assumed distribution, the ratio changes over time (\fref{Rtbias}F,I).
In particular, if the true generation intervals have a shorter (or longer) mean than the assumed distribution, we over-estimate (or under-estimate) $\Ry{var}(t)$ during the growth phase and under-estimate (or over-estimate) $ $ during the decay phase (\fref{Rtbias}D,G), which further translates to biases in the estimated relative strength (\fref{Rtbias}F,I).

We can assess these biases by plotting the estimated strength of the variant $\hat{\RR}_{\textrm{var}}(t)$ against the estimated strength of the wild type $\hat{\RR}_{\textrm{wt}}(t)$ on a log-log scale (\fref{Rtbias}C,F,I).
When generation-interval distributions are correctly specified, we obtain a straight line with a slope of 1. 
In this case, the intercept corresponds to the logarithm of the relative strength $\log(\rho)$ (\fref{Rtbias}C).
When the assumed mean generation interval is longer (or shorter) than the that of the variant, we over-estimate (or under-estimate) the slope (\fref{Rtbias}H,L) on a log-log plot.
Smooth changes in $\RR$ further generate circular patterns but biases in slopes are clear.

\section{Implications for intervention strategies}

While the relative speed need not stay constant over the course of an epidemic, at any given point in epidemic, we can measure the speed of the new variant and the wild type and ask how much more intervention is required to control the spread of both strains (relative to measured conditions).
As a baseline scenario, we assume $\ry{wt}=0\,\textrm{/days}$ and $\delta=0.1\,\textrm{/days}$ (and therfore $\ry{var}=0\,\textrm{/days}$).
In this case, the wild type is already under control, whereas the variant requires additional intervention measures to stop growing.

Here, we consider two types of interventions:
a strength-like intervention which affects the kernel by a constant amount $\theta$ regardless of when they were infected ($K_{\mathrm{post}}(\tau) = K_{\mathrm{pre}}(\tau)/\theta$) and a speed-like intervention which affect the kernel at a constant rate $\phi$ ($K_{\mathrm{post}}(\tau) = K_{\mathrm{pre}}(\tau) \exp(-\phi \tau)$).
In this case, it is clear that we can control the spread of the variant when either $\theta > \Ry{var}$ or $\phi > \ry{var}$---these two quantities ($\theta$ and $\phi$) are also referred to as strength and speed of intervention, respectively.
In general, we can calculate the strength ($\theta = \Ry{pre}/\Ry{post}$) and speed ($\phi = \ry{pre} - \ry{post}$) of intervention, both providing equivalent measures for effectiveness of intervention, for any intervention by comparing epidemiological conditions before and after the intervention \citep{doi:10.1098/rspb.2020.1556}---these quantities, by definition, must be greater than pre-intervention strength and speed of the epidemic to control the spread.
We compare how potential differences in generation-inteval distributions affect the control of the spread of the new variant using this strength-and-speed framework.

\begin{figure}[!th]
\includegraphics[width=\textwidth]{control.pdf}
\caption{
\textbf{Strength and speed of intervention against the new variant.}
(A) Strength of epidemic (black, solid), constant strength intervention (orange, dashed), and constant speed intervention (blue, dotted).
(B) Speed of epidemic (black, solid), constant strength intervention (orange, dashed), and constant speed intervention (blue, dotted).
Strength and speed are calculated assuming $\ry{wt}=0\,\textrm{/days}$, $\delta=0.1\,\textrm{/days}$, $\ry{var}=0\,\textrm{/days}$, $\Gy{wt}=5\,\textrm{days}$, and $\kappa=1/5$.
Constant strength and speed interventions are modeled such that post intervention strength of the new variant, assuming equal mean generation intervals at 5 days, corresponds to 0.9.
}
\label{fig:strengthspeed}
\end{figure}

\fref{strengthspeed} compares the effectiveness of strength-based and speed-based interventions across a wide range of assumptions about generation-interval distributions.
We begin by assuming that the variant and the wild type have equal mean generation intervals and model interventions such that they both reduce $\Ry{var}$ to 0.9 after intervention (\fref{strengthspeed}A).
If the mean generation interval of the variant is longer than that of the wild type, then the strength of the variant will be higher;
since the strength of the strength-like intervention is insensitive to the underlying generation-interval distribution, we may fail to control the spread of the new variant.
The strength of the speed-like intervention, however, depends on the new variant:
since speed-like interventions, such as contact tracing, have disproportionate effects on preventing later transmission, they are more effective (therefore stronger) when the mean generation interval of the new variant is longer.

The speed-based paradigm gives same results about the control but provides important insights (\fref{strengthspeed}B).
The observed speed of the variant at any given moment does not depend on our estimates of its mean generation interval.
Therefore, if we can prevent infections at a faster rate than the observed speed of spread (i.e., if the speed of intervention is faster than the epidemic speed), we can control the epidemic---this explains why a constant-speed intervention can prevent the spread of the variant regardless of its mean generation interval.
On the other hand, differences in mean generation intervals between the variant and the wild type imply different post-intervention epidemic speed (therefore, intervention speed) under a constant-strength intervention, explaining why it might not be effective if the variant has longer generation intervals.

@@

Here, we explored how the generation-interval distribution shapes the link between relative strength and speed of the new variant.
We define relative strength as the ratio between reproduction numbers ($\rho=\Ry{var}/\Ry{wt}$) and relative speed as the difference between growth rates ($\delta=\ry{var}-\ry{wt}$) of the variant and the wild type.
When the generation-interval distributions are assumed to be identical, we find that estimates of $\hat{\rho}$ decreases with $r_w$ whereas $\hat{\delta}$ increase with $\RR_w$.
When we account for possible differences in generation-interval distributions, these relationships can change directions and further lead to biases in these estimates.
We also show that differences in generation intervals lead to different conclusions about the effectiveness of interventions:
if the variant has longer or shorter generation intervals than the wild type, we should prioritize speed- or strength-like interventions, respectively.

We argue that both perspectives (i.e., whether we hold relative strength or speed constant) are useful in understanding the dynamics of the new variant.
Early in the spread, we typically want to fix the relative speed because we can directly observe it.
As more information about the transmission and immunity profiles of the new variant becomes available, it is better to fix the relative strength as it better matches biological mechanisms for its higher strength.
We suggest researchers to be mindful about which quantity they hold constant and how their assumptions lead to their conclusions.

>>>>

\newcommand{\rx}[1]{\ensuremath{{r}_{#1}}\xspace} 
\newcommand{\ry}[1]{\rx{\mathrm{#1}}}
\newcommand{\rw}{\ry{wt}}

\newcommand{\Rx}[1]{\ensuremath{{\mathcal R}_{#1}}\xspace} 
\newcommand{\Ry}[1]{\Rx{\mathrm{#1}}}
\newcommand{\Ro}{\Rx{0}}
\newcommand{\Rc}{\Rx{\mathrm{c}}}
\newcommand{\Ri}{\Rx{\mathrm{i}}}
\newcommand{\RR}{\ensuremath{{\mathcal R}}\xspace}
\newcommand{\Rhat}{\ensuremath{{\hat\RR}}}
\newcommand{\Rnaive}{\ensuremath{{\mathcal R}_{\textrm{\tiny naive}}}\xspace}
\newcommand{\tsub}[2]{#1_{{\textrm{\tiny #2}}}}
\newcommand{\days}{\ensuremath{\, \textrm{day}}}
\newcommand{\pday}{\ensuremath{/\textrm{day}}}
\newcommand{\dd}[1]{\ensuremath{\, \mathrm{d}#1}}
\newcommand{\dtau}{\dd{\tau}}
\newcommand{\dx}{\dd{x}}
\newcommand{\dsigma}{\dd{\sigma}}

@@

Epidemic speed $r_x$ can be often estimated directly from case data \citep{mills2004transmissibility,nishiura2009transmission,ma2014estimating};
this can be challenging when case counts are low (as when a new variant begins to spread) and is sensitive to temporal changes in testing patterns and intensity.
Studies of new SARS-CoV-2 variants have mostly focused on characterizing changes in \emph{proportion} of a new variant \citep{switzerland2021variant, davies2021estimated, di2021impact, leung2021early, volz2021transmission,zhao2021}.
This has the advantage of being less sensitive to changes in testing and to other transient effects that might affect variants and wild type viruses similarly.
When incidence is changing exponentially ($i_x(t) = i_x(t_0) \exp(r_x t)$), the proportion of the new variant $p(t)$ follows a logistic equation:
\begin{align}
p(t) &= \frac{\iy{var}(t_0) \exp(\ry{var} t)}{\iy{wt}(t_0) \exp(\ry{wt} t) + \iy{var}(t_0) \exp(\ry{var} t)},
\\ &= \frac{1}{1 + \left(\iy{wt}(t_0)/\iy{var}(t_0)\right) \exp(-\delta t)},
\end{align}
where the logistic growth rate $\delta = \ry{var} - \ry{wt}$ corresponds to the relative speed of the epidemic.

\begin{figure}[!th]
\includegraphics[width=\textwidth]{relstrength.pdf}
\caption{
\textbf{Relative strength of the new variant given observed relative speed.}
(A) Estimated relative strength of the new variant $\hat{\rho}$ given speed of the wild type
$\ry{wt}$ assuming identical generation-interval distributions.
(B) True relative strength of the new variant $\rho$ given speed of the wild type $\ry{wt}$ and ratio between the mean generation interval of the new variant $\bar{G}_v$ and that of the wild type $\bar{G}_w$.
(C) Ratio between the true and the estimated relative strength of the new variant $\rho/\hat{\rho}$.
True relative strength of the new variant $\rho$ is calculated using $\delta=0.1\pday$, $\bar{G}_w = 5\days$, and $\kappa = 1/5$.
}
\label{fig:relstrength}
\end{figure}

We thus ask: what factors affect the relative strength $\rho = \Ry{var}/\Ry{wt}$ of a new variant, given an observed relative speed $\delta$?
This inference will depend in general on assumptions about the growth rate of the wild type and on the generation-interval distributions of both variants.
These have generally been assumed to be the same, but we will also consider the possibility that they are different.

As our baseline scenario, we take the relative speed of the variant to be $\delta = 0.1/\textrm{day}$ \citep{davies2021estimated}; the mean generation interval of the wild type to be $\bar{G}_w = 5\days$ \citep{ferretti2020quantifying}; and the squared coefficient of variation of generation intervals to be $\kappa=0.2$ \citep{ferretti2020quantifying}.
For simplicity, we assume that the variant and the wild type have equal $\kappa$ throughout, and only consider differences in the mean.
\jd{This is awkward; let's skip for now:} As we expect there to be greater uncertainty in $\kappa$ than in $\bar{G}$, we evaluate the estimates of relative strength $\rho$ across a wide range of $\kappa$, ranging from 0 (delta distribution) to 1 (exponential distribution).

When the variant and wild type have the same generation-interval distributions, $\Gy{var} = \Gy{wt} = \bar{G}$, 
the estimate of the relative strength is given by \citep{park2019practical}:
\begin{equation}
\hat{\rho} = \left(1 + \frac{\kappa \delta \bar{G}}{1 + \kappa \ry{wt} \bar{G}}\right)^{1/\kappa}.
\label{eq:rhoGbar}
\end{equation}
Therefore, estimates of relative strength $\rho$ depend not only on the relative speed $\delta$ but also on how fast the wild type is spreading in the population \rw---a few analyses have implicitly or explicitly neglected this factor by either assuming $\rw = 0$ \citep{switzerland2021variant} or $\kappa = 0$ \citep{davies2021estimated} (in the latter case $\hat{\rho} = \exp(\delta \bar{G})$).

\eref{rhoGbar} shows that, for realistic (non-zero) values of $\kappa$, estimates of the strength $\rho$ go down as estimates of the growth rate $\rw$ go up. 
\jd{I like effective dispersion, but it does no good to just mention it once. Leave it out for now?}
The relative strength also depends on the \emph{effective dispersion} in generation intervals (i.e., the amount of variability in generation intervals, as measured by $\kappa$)---using narrower distributions makes $\hat{\rho}$ less sensitive to $\ry{wt}$.
These effects are illustrated in \fref{relstrength}A.

However, if the true mean generation interval of the new variant is different from that of the wild type ($\bar{G}_v \neq \bar{G}_w$), the true relative strength $\rho = \RR_v/\RR_w$ is given by:
\begin{equation}
\rho = \left(\frac{1 + \kappa (\rw + \delta) \bar{G}_v}{1 + \kappa \rw \bar{G}_w}\right)^{1/\kappa}.
\end{equation}
In this case, longer (or shorter) mean generation interval of the new variant generally translates to higher (or lower) values of $\rho$---\fref{relstrength}B illustrates this effect.
However, this direction can change when both the new variant and the wild type are decaying exponentially (see area below $r_v = 0$ line in \fref{relstrength}B)---for a given rate of decline, longer generation intervals lead to a \emph{lower} value of strength (and therefore lower relative strength $\rho$).

Therefore, neglecting potential differences in generation intervals can lead to mis-estimation of the relative strength and the variant, with implications for forecasting and planning.
In \fref{relstrength}C, we quantify the bias that arises by incorrectly assuming generation intervals are the same using the ratio between the true value $\rho$ and the estimated value $\hat{\rho}$, which assumes equal generation-interval distributions.
Across the ranges of parameters that we consider, we find that up to 1.4 fold (either upward or downward) adjustments to current estimates may be required.

@@

While relative speed is easier to measure, we do not generally expect $\delta$ to be constant if other factors governing epidemic spread are changing.
Instead, many biological mechanisms better translate to assuming a constant value of relative strength $\rho$ over changing conditions.
For example, if the new variant has a higher transmission rate, the ratio between the strength of the new variant and that of the wild type will remain constant ($=\rho$) under strength-like interventions that reduces the strength of both strains by a constant amount.
However, this is not necessarily true under speed-like interventions (e.g., contact tracing and self-isolation), which depends on time since infection: if the generation-interval distributions are different, the effectiveness of interventions will also differ.
For now, we consider consequences of assuming a constant relative strength $\rho$ and predicting the relative speed $\delta$---an assumption often made (either implicitly or explicitly) in studies that have tried to predict the future spread of the new variant, including when it will replace the wild type.
As our baseline scenario we assume $\rho = 1.5$.

\begin{figure}[!th]
\includegraphics[width=\textwidth]{relspeed.pdf}
\caption{
\textbf{Relative speed of the new variant given assumed relative strength.}
(A) Estimated relative speed of the new variant $\hat{\delta}$ given strength of the wild type $\RR_w$ assuming identical generation-interval distributions.
(B) True relative strength of the new variant $\delta$ given strength of the wild type $\RR_w$ and ratio between the mean generation interval of the new variant $\bar{G}_v$ and that of the wild type $\bar{G}_w$..
(C) Difference between the true and the estimated relative speed of the new variant $\delta - \hat{\delta}$.
True relative speed of the new variant $\delta$ is calculated using $\rho=1.5$, $\bar{G}_w = 5\,\textrm{days}$, and $\kappa=1/5$.
}
\label{fig:relspeed}
\end{figure}

@@

In practice, epidemiological conditions change over time due to intervention measures and susceptible depletion effects.
Such changes can be characterized by ``time-varying'' or ``instantaneous'' reproduction numbers, $\RR(t)$ \citep{fraser2007estimating}.
Assuming that the generation-interval distribution does not change over time, the instantaneous reproduction numbers of the new variant and of the wild type can be estimated from their corresponding incidence curves---an approach popularized by \cite{cori2013new}:
\begin{equation}
\RR_x(t) = \frac{i_x(t)}{\int_0^\infty i_x(t-\tau) g_x(\tau) \dtau}.
\end{equation}
Under strength-like intervention measures that reduce transmission rates by a constant amount, we expect ratios between true reproduction numbers to remain constant and correspond to the true relative strength: $\RR_v(t)/\RR_w(t) = \rho$.
However, if the assumed generation-interval distribution $\hat{g}(\tau)$ is different from the true distribution, then the ratio between the estimated reproduction numbers $\hat{\rho}(t) = \hat{\RR}_v(t)/\hat{\RR}_w(t)$ need not be constant, where
\begin{equation}
\hat{\RR}_x(t) = \frac{i_x(t)}{\int_0^\infty i_x(t-\tau) \hat{g}(\tau) \dtau}.
\end{equation}

Here, we simulate three different scenarios using a two strain renewal equation that assumes perfect cross-immunity (\fref{Rtbias}): 
(1) the wild type and the variant have the same generation-interval distributions, which match the assumed distribution (\fref{Rtbias}A--D);
(2) the assumed distribution matches the generation-interval distribution of the wild type but has a longer mean than the mean generation interval of the variant (\fref{Rtbias}E--H); and
(3) the assumed distribution matches the generation-interval distribution of the wild type but has a shorter mean than the mean generation interval of the variant (\fref{Rtbias}I--L).
Then, we compare the estimated ratio $\hat{\rho}(t) = \hat{\RR}_v(t)/\hat{\RR}_w(t)$ with the true ratio $\rho = \RR_v(t)/\RR_w(t)$.
For simplicity, we did not model non-pharmaceutical interventions;
instead, we let epidemic dynamics to be determined by susceptible depletion alone.

\begin{figure}[!th]
\includegraphics[width=\textwidth]{Rtbias.pdf}
\caption{
\textbf{Estimates of relative strength over time under different scenarios.}
(A, E, I) Daily incidence of infection caused by the new variant and the wild type.
(B, F, J) True (solid lines) and estimated (dashed lines) reproduction numbers of the new variant and the wild type over time.
(C, G, K) True (purple) and estimated (orange) ratios between reproduction numbers of the new variant and the wild type over time.
(D, H, L) Relationship between estimated reproduction numbers on a log-log sacle.
The dashed lines represent the one-to-one line. 
The dotted lines represent the $\RR_v=\rho \RR_w$ line. 
Top row: the assumed mean generation-interval (5 days) is equal to the mean generation-interval of the wild type and the variant.
Middle row: the assumed mean generation-interval (5 days) is equal to the mean generation-interval of the wild type but is longer than that of the varant (3.5 days).
Bottom row: the assumed mean generation-interval (5 days) is equal to the mean generation-interval of the wild type but is longer than that of the varant (6.5 days).
For all simulations, squared coefficient of variation in generation intervals is assumed to equal $\kappa = 1/5$.
}
\label{fig:Rtbias}
\end{figure}

When the assumed distribution matches the true distributions, the estimated ratio remains constant (\fref{Rtbias}C).
However, if the generation-interval distribution of the variant is different from the assumed distribution, the ratio changes over time (\fref{Rtbias}G,K).
In particular, if the true generation intervals have a shorter (or longer) mean than the assumed distribution, we over-estimate (or under-estimate) $\RR_v(t)$ during the growth phase and under-estimate (or over-estimate) $\RR_v(t)$ during the decay phase (\fref{Rtbias}F,J), which further translates to biases in the estimated relative strength (\fref{Rtbias}G,K).

We can also assess these biases by plotting the estimated strength of the variant $\hat{\RR}_v(t)$ against the estimated strength of the wild type $\hat{\RR}_w(t)$ on a log-log scale (\fref{Rtbias}D,H,L).
When generation-interval distributions are correctly specified, we obtain a straight line parallel to the one-to-one line. 
In this case, the slope corresponds to the logarithm of the relative strength $\log(\rho)$ (\fref{Rtbias}D).
When the assumed mean generation interval is longer (or shorter) than the that of the variant, we over-estimate (or under-estimate) the slope (\fref{Rtbias}H,L) on a log-log plot.

\section{Inferring relative strength from case data, in practice}

Seb Funk's analysis here.

\section{Intervention}

If $K(\tau)/L(\tau)$.

@@

Here, we explored how the generation-interval distribution shapes the link between relative strength and speed of the new variant.
We define relative strength as the ratio between reproduction numbers ($\rho=\RR_v/\RR_w$) and relative speed as the difference between growth rates ($\delta=r_v-\rw$) of the variant and the wild type.
When the generation-interval distributions are assumed to be identical, we find that estimates of $\hat{\rho}$ decreases with $\rw$ whereas $\hat{\delta}$ increase with $\RR_w$.
When we account for possible differences in generation-interval distributions, these relationships can change directions and further lead to biases in these estimates.

We argue that both perspectives (i.e., whether we hold relative strength or speed constant) are useful in understanding the dynamics of the new variant.
Early in the spread, we typically want to fix the relative speed because we can directly observe it.
As more information about the transmission and immunity profiles of the new variant becomes available, it is better to fix the relative strength as it better matches biological mechanisms of its spread.
We suggest researchers to be mindful about which quantity they hold constant and how their assumptions lead to their conclusions.
